---
title: "Loop Assignment"
author: "Friedrich Knuth"
date: "February 11, 2015"
output: html_document
---

```{r}
data(iris)
head(iris)
sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) = sp_ids
colnames(output) = names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        x = 0
        y = 0
        if (nrow(iris_sp) > 0) {
            for(k in 1:nrow(iris_sp)) {
                x = x + iris_sp[k, j]
                y = y + 1
            }
            output[i, j] = x / y 
        }
    }
}
output
```

**Q1. Describe the values stored in the object `output`. In other words what did the loops create?**

```{r}
iris_sp = subset(iris, Species == "setosa")
mean(iris_sp$Sepal.Length)
```

*The preceeding code shows us that the mean `Sepal.Length` for `setosa` is 5.006. It seems that the for loop generated an output reflecting the mean value for `Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width`.*

**Q2. Describe using pseudo-code how output was calculated.**

- For every species  
    - For every physical descriptor  
        - For every observation  
            - Add the value sequentially for every observation to x (x starts at 0)  
            - Add 1 sequentially for every observation to y (y starts at 0)  
        
Within the `output` matrix, for every row entry in sequence along `sp_ids` (ie for every species) and in columns from `iris_sp` (ie every physical descriptor) devide the final sum of physical descriptor values (x) by the final
sum of all observations (y). This results is the mean value for each physical descriptor by species, represented by the matrix `output`.
        
**Q3. The variables in the loop were named so as to be vague. How can the objects `output`, `x`, and `y` could be renamed such that it is clearer what is occurring in the loop.**

`output` = species_mean  
`x` = sample_sum  
`y` = samples_count  

**Q4. It is possible to accomplish the same task using fewer lines of code? Please suggest one other way to calculate output that decreases the number of loops by 1.**

`if (nrow(iris_sp) > 0) {}` seems unnecessary. The code produces the same result without that loop. I am not sure what it is for. 

```{r}
data(iris)
head(iris)
sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) = sp_ids
colnames(output) = names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        x = 0
        y = 0
        for(k in 1:nrow(iris_sp)) {
            x = x + iris_sp[k, j]
            y = y + 1
            }
            output[i, j] = x / y 
    }
}
output
```
